Synthesizing Symmetric Lenses

### 2019-02-28 papers/SymmetricLenses/main.tex
Finished with polishing 1 and 2. Lots of comments. Switching to other papers now, but I will be available for discussion.

The expected number of bits required to recover a piece of data corresponds to
the well-known information-theoretic concept of
\emph{entropy}~\cite{Shannon1948}. Calculating entropy requires a probability
distribution over the space of possible values for the data. Specifically, given
a set $S$ and a probability distribution $P \OfType S \rightarrow \Reals$ over
$S$, the entropy $\EntropyOf{S,P}$ is $-\Sigma_{s\in S}P(s)\Log_2P(s)$. In
information theory, each element of $s \in S$ has \emph{information content},
the number of bits required to specify that element in a perfect encoding scheme
($-\Log_2P(s)$). The entropy of $S$ calculates the expected number of bits
required to specify an element drawn from $S$, in other words, the
probability of each element times its information content. Entropy maintains
that if a data source contains many possible elements, where none have
significantly higher probability than others, it will have high information
content. Data sources with few elements that don't have vanishingly small
probabilities have small entropy. When $P$ is clear, we will often use the
shorthand of $\EntropyOf{S}$ for $\EntropyOf{S,P}$.

In our setting, we already have a way of expressing sets of data: regular
expressions. To calculate entropy, what we need is a way to express probability
distributions over that data. To that end, we adopt \emph{stochastic regular
  expressions}~\cite{stoch-def,stoch-rnn} (SREs), which are regular expressions
in which each operator is annotated with a probability (see \S\ref{sec:sre}). A
stochastic regular expression specifies both a set of strings and a probability
distribution over that set.

We use entropy to guage the relative likelihood of such lenses in our synthesis
algorithm (see \S\ref{subsec:lenscosts}). For any lens $\Lens$, we can calculate
the expected number of bits required to recover a string $t$ in
$\LanguageOf{\RegexAlt}$ from a synchronized string $s$ in
$\LanguageOf{\Regex}$. This expectation is the \emph{conditional entropy of
  $\Regex$ given $\RegexAlt$ and $\Lens$}, formally $\sum_{s \in
  \Regex}\ProbabilityOf{\Regex}{s}\cdot\EntropyOf{\SetOf{t \SuchThat
    \Lens.\PutROf{s}{t} = t}}$. The likelihood we assign to $\Lens$ is the sum
of the conditional entropy of $\Regex$ given $\RegexAlt$ and $\Lens$ and the
conditional entropy of $\RegexAlt$ given $\Regex$ and $\Lens$. This metric
assigns higher entropy (or cost) to lenses where knowing the string on one side
provides little information about the string on the other side. It assigns zero
entropy to bijections because given a string $s \in \Regex$, the bijection
exactly determines the corresponding string in $\RegexAlt$.

@@

To obtain SREs from the plain regular expressions that users write, by default
we use a heuristic that attempts to assign probability annotations that give
each string in the language equal probablity. In other words, this heuristic
tries to make each bit of data no more or less likely than any other.
Our synthesis algorithm could equally well take as input a pair of stochastic
regular expression rather than a plain one, permitting users to specify a
data-specific probability distribution (either manually calculated or inferred
from some dataset of interest).

However, sometimes users already know that certain data should or should not be
used to construct the other format. We introduce relevance annotations to SREs
that enable users to specify whether a piece of data should be used to construct
the other format (with \SRequire) or not (with \Skip). In our example,
\lstinline{salary} and \lstinline{company} could safely be skipped (as they are
disconnected), where \lstinline{name} in \lstinline{emp_salary} and
\lstinline{emp_ins} be annotated as required (as they are converted with the
identity lens).

\subsection{Searching for Likely Lenses}
Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work~\cite{augustsson-2004, osera+:pldi15,
  feser-pldi-2015, frankle+:popl16} has shown that we can use types to
dramatically restrict the search space and improve the effectiveness of
synthesis. In our setting, types are pairs of stochastic regular
expressions. The fact that each regular expression is semantically equivalent to
infinitely many others complicates the problem.
Following existing work on bidirectional program
synthesis~\cite{maina+:quotient-synthesis}, we 
split our algorithm into two communicating
searches. The first search,
\Expand, navigates the space of semantically equivalent regular expressions
by applying rewrite rules that preserve both the semantics and the
probability distributions of those stochastic regular expressions. This
algorithm ranks pairs of stochastic regular expressions by the number of
rewrite rule applications required to obtain each pair from the one given as
input. It passes the pairs off to the second 
search procedure, \GreedySynth, in rank order with the smallest first.

Abstractly, \GreedySynth looks for highest-likelihood lenses between a given
pair of stochastic regular expressions $\Regex$ and $\RegexAlt$ by performing a
type-directed search. In more detail, \GreedySynth first converts the stochastic
regular expressions provided by \Expand into \emph{stochastic DNF regular
  expressions}---a constrained representation of stochastic regular expressions
with disjunctions distributed over concatenations and with concatenations and
disjunctions normalized to operating over lists (see \S\ref{subsec:sdnfre}).
\GreedySynth then uses the syntax of these n-ary DNF regular expressions to find
normalized lenses, in a form we call \emph{simple symmetric n-ary DNF lenses}.
These lenses include neither a composition operator nor a type equivalence rule.
These restrictions mean that there are only a few simple symmetric n-ary DNF
lenses that are well-typed between each pair of stochastic n-ary DNF regular
expressions, hence \GreedySynth's search space for a given pair of regular
expressions is finite. Finally, \GreedySynth yields a simple symmetric lens by
converting the n-ary forms into the binary forms provided in the surface
language.

This architecture of communicating synthesizers gives us a way to enumerate
pairs of stochastic regular expressions of increasing rank and to efficiently
search through them, but it poses a problem: when should \Expand stop proposing
new SRE pairs? We may have found a promising lens between a pair of stochastic
regular expressions, but a different pair we haven't yet discovered may give
rise to an even better lens. The search algorithm must resolve a tension between
the quality of the inferred lens and the amount of time it takes to return a
result. For example, if the algorithm has already found the lens in
Figure~\ref{fig:example_lens}, we don't want to spend a lot of time searching
for an even better lens. To resolve this tension, the algorithm uses heuristics
to judge whether to return the current best satisfying lens to the user or to
pass the next set of unrollings to \GreedySynth. The heuristics favor stopping
if the current best satisfying lens is very likely, indicating the lens is very
promising (for example, if the satisfying lens loses no information, the
algorithm should terminate for no other lens will be more likely). The
heuristics also favor stopping if \Expand has delivered to \GreedySynth all
pairs of stochastic regular expressions at a given rank and there is a large
number of pairs at the next rank, because searching through all such pairs will
take a long time. With this approach, the algorithm can quickly respond with a
satisfying lens with relatively high likelihood. If the user is unhappy with the
result, they can either refine the search by supplying additional examples,
which serve both to rule out the previously proposed lens(es) and to reduce the
size of the search space by cutting down on the number of satisfying lenses, or
they can supply annotations on the source and target SREs indicating that
certain information either must be retained or must be discarded by the lens
(see \S\ref{subsec:optimizations}).

>>>>

The expected number of bits required to recover a piece of data corresponds to
the well-known information-theoretic concept of
\emph{entropy}~\cite{Shannon1948}. Calculating entropy requires a probability
distribution over the space of possible values for the data. Specifically, given
a set $S$ and a probability distribution $P \OfType S \rightarrow \Reals$ over
$S$, the entropy $\EntropyOf{S,P}$ is $-\Sigma_{s\in S}P(s)\Log_2P(s)$. In
information theory, the \emph{information content} of each element of $s \in
S$ is 
the number of bits required to specify $s$ in a perfect encoding scheme
($-\Log_2P(s)$). The entropy of $S$ is the expected number of bits
required to specify an element drawn from $S$---the
probability of each element times its information content. Entropy captures
the intuition that, if a data source contains many possible elements and
none have 
significantly higher probability than others, it will have high information
content. Data sources where most elements have small
probabilities have small entropy. When $P$ is clear from context, we will
often use the shorthand of $\EntropyOf{S}$ for $\EntropyOf{S,P}$.

In the present setting, we already have a way of expressing sets of data:
regular 
expressions. To calculate entropy, what we need is a way to express probability
distributions over those sets. To that end, we adopt \emph{stochastic regular
  expressions}~\cite{stoch-def,stoch-rnn} (SREs), which are regular expressions
in which each operator is annotated with a probability (see \S\ref{sec:sre}). A
stochastic regular expression thus specifies both a set of strings at the
same time as a probability distribution over that set.

We use entropy to gauge the relative likelihood of lenses in our synthesis
algorithm (see \S\ref{subsec:lenscosts}). For any lens $\Lens$, we can calculate
the expected number of bits required to recover a string $t$ in
$\LanguageOf{\RegexAlt}$ from a synchronized string $s$ in
$\LanguageOf{\Regex}$. This expectation is the \emph{conditional entropy of
  $\Regex$ given $\RegexAlt$ and $\Lens$}, formally $\sum_{s \in
  \Regex}\ProbabilityOf{\Regex}{s}\cdot\EntropyOf{\SetOf{t \SuchThat
    \Lens.\PutROf{s}{t} = t}}$. The likelihood we assign to $\Lens$ is the sum
of the conditional entropy of $\Regex$ given $\RegexAlt$ and $\Lens$ and the
conditional entropy of $\RegexAlt$ given $\Regex$ and $\Lens$. This metric
assigns higher entropy (or cost\bcp{= lower likelihood?  I'm still confused
  by these different ways of speaking about what seems to be the same
  thing.}) to lenses where knowing the string on one side 
provides little information about the string on the other side. It assigns zero
entropy to bijections because given a string $s \in \Regex$, the bijection
exactly determines the corresponding string in $\RegexAlt$.

@@

To obtain SREs from the plain regular expressions that users write, 
we use a default heuristic that attempts to assign probability annotations
giving
each string in the language equal probablity. In other words, this heuristic
tries to make each bit of data no more or less likely than any
other.\bcp{huh?} 
Alternatively, our synthesis algorithm could take as input a pair of stochastic
regular expressions rather than plain ones, permitting users to specify a
data-specific probability distribution (either manually calculated or inferred
from some dataset of interest).\bcp{So why didn't we do it this way?}

However\bcp{``However'' means that something from before is being
  contradicted.  What is it in this case?}, sometimes users already know that certain data should or should not be
used to construct the other format. We introduce relevance annotations to SREs
that enable users to specify whether a piece of data should be used to construct
the other format (with \SRequire) or not (with \Skip). In our example,
\lstinline{salary} and \lstinline{company} could safely be skipped (as they are
disconnected), where \lstinline{name} in \lstinline{emp_salary} and
\lstinline{emp_ins} be annotated as required (as they are converted with the
identity lens).  \bcp{That paragraph seems way too brief and high level.
  These annotations were added in response to significant complaints from an
earlier PC, but the paper reads as though they are an afterthought of little
consequence; they deserve more attention and explanation.}

\subsection{Searching for Likely Lenses}
Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work~\cite{augustsson-2004, osera+:pldi15,
  feser-pldi-2015, frankle+:popl16} has shown that types can be used to
dramatically restrict the search space and improve the effectiveness of
synthesis. In our setting, types are pairs of stochastic regular
expressions, and the fact that each regular expression is semantically
equivalent to infinitely many others complicates the problem.

Following existing work on bidirectional program
synthesis~\cite{maina+:quotient-synthesis}, we 
split our algorithm into two communicating
searche procedures. The first,
\Expand, navigates the space of semantically equivalent regular expressions
by applying rewrite rules that preserve both semantics and
probability distributions. This
algorithm ranks pairs of stochastic regular expressions by the number of
rewrite rule applications required to obtain each pair from the one given as
input. It passes the pairs off to the second 
search procedure, \GreedySynth, in rank order with the smallest first.

\GreedySynth looks for highest-likelihood lenses between a given
pair of stochastic regular expressions $\Regex$ and $\RegexAlt$ by performing a
type-directed search. It first converts the stochastic
regular expressions provided by \Expand into \emph{stochastic DNF regular
  expressions}---a constrained representation of stochastic regular expressions
with disjunctions distributed over concatenations and with concatenations and
disjunctions normalized to operating over lists.
Then it uses the syntax of these n-ary DNF regular expressions to find
normalized lenses in a form we call \emph{simple symmetric n-ary DNF lenses}.
These involve neither a composition operator nor a type equivalence rule.
These restrictions mean that there are only a few simple symmetric n-ary DNF
lenses that are well typed between a given pair of stochastic n-ary DNF regular
expressions, so \GreedySynth's search space for a given pair of regular
expressions is finite. Finally, \GreedySynth yields a simple symmetric lens by
converting the n-ary syntax back to the binary forms provided in the surface
language.

This architecture of communicating synthesizers gives us a way to enumerate
pairs of stochastic regular expressions of increasing rank and to efficiently
search through them, but it poses a problem: when should \Expand stop proposing
new SRE pairs? We might have found a promising lens between a pair of stochastic
regular expressions, but a different pair we haven't yet discovered may give
rise to an even better lens. The search algorithm must resolve a tension between
the quality of the inferred lens and the amount of time it takes to return a
result. For example, if the algorithm has already found the lens in
Figure~\ref{fig:example_lens}, we don't want to spend a lot of time searching
for an even better lens. To resolve this tension, the algorithm uses heuristics
to judge whether to return the current best satisfying lens to the user or to
pass the next set of unrollings to \GreedySynth. The heuristics favor stopping
if the current best satisfying lens is very likely, indicating the lens is very
promising (for example, if the satisfying lens loses no information, the
algorithm should terminate for no other lens will be more likely\bcp{this
  case is 100\% obvious, which makes it not so interesting as an example.
  The reader is left wondering what happens with pretty likely but not
  maximally likely lenses?}). The
heuristics also favor stopping if \Expand has delivered to \GreedySynth all
pairs of stochastic regular expressions at a given rank and there is a large
number of pairs at the next rank, because searching through all such pairs will
take a long time. With this approach, the algorithm can quickly respond with a
satisfying lens with relatively high likelihood. If the user is unhappy with the
result, they can either refine the search by supplying additional examples,
which serve both to rule out previously proposed lenses and to reduce the
size of the search space by cutting down on the number of satisfying lenses, or
they can supply annotations on the source and target SREs indicating that
certain information either must be retained or must be discarded by the lens
(see \S\ref{subsec:optimizations}). \bcp{Again, I feel like we're
  downplaying this aspect too much.}
